The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2020-03-24 razzo-abstract.tex
Feedback rsetienne

% One or two sentences providing a basic
% introduction to the field,
% comprehensible to a scientist in any discipline.
Millions of species exist on Earth, all originating from a common ancestor billions
of years ago.
The field of phylogenetics uses heritable material (e.g. DNA) to determine the evolutionary history of species.

% Two to three sentences of
% more detailed background, comprehensible to
% scientists in related disciplines.
Starting from heritable material and explicit assumptions, Bayesian phylogenetics allows to infer a jointly-estimated phylogeny and parameter estimates distribution.
One of these assumptions is the speciation model, which mathematically describes how species are added or removed from a phylogeny in time.
Speciation models commonly assume that speciation events are independent and disjointed. Yet, such assumption may overlook the complexity of certain processes, where environmental changes promote speciation events in multiple lineages.
% Purpose of the paper
This new layer of complexity can be captured developing a novel ad hoc speciation model. However, the introduction of a new model could be not necessary if current models are already capable of describing the process to a satisfactory extent, under simpler assumptions.
Here we investigate the extent of the discrepancy produced by current BEAST2's tree priors, when trying to infer back trees where speciation can co-occur.

% Summarising the main result
We let the impact of co-occurring speciation vary on datasets of simulated trees and show the corresponding error produced during the inference process.

% Two or three sentences explaining what
% the main result reveals in direct
% comparison to what was thought to be the case
% previously, or how the main result adds to
% previous knowledge.
We show that the extent of the inference error increases with the amount of co-occuring speciation events to establish the limits of BEAST2's standard tree priors in this circumstance.

% One or two sentences to put the results into a
% more general context.
These results allow phylogeneticists to judge under which circumstances the commonly used speciation model can be safely used. \giovanni{Not entirely sure if this is the best sentence to put here.}

>>>>

% One or two sentences providing a basic
% introduction to the field,
% comprehensible to a scientist in any discipline.
The field of phylogenetics uses heritable material such as DNA to determine the (shared) evolutionary history of a set of species which are summarized in a phylogenetic tree.
% Two to three sentences of
% more detailed background, comprehensible to
% scientists in related disciplines.
Bayesian phylogenetic methods allow us to jointly infer probability distributions for the phylogenetic tree and parameters of the various models underlying the methods.
One of these models is the diversification model, which mathematically describes the dynamics of speciation addition and removal through time. In Bayesian analyses these are called the (species) tree priors. Tree priors commonly assume that speciation events occur independently. The Bayesian tools heavily rely on this assumption. However, under species pump dynamics, this assumption is violated. By species pump dynamics we mean the (repeated) simultaneous formation of multiple species due to environment-driven isolation which results in temporally aligned (or clustered) divergence times.
% Purpose of the paper
Current Bayesian phylogenetic tools do not contain such species pump diversification models as tree priors. This may not be a problem if currently implemented tree priors are already capable of inferring a phylogenetic tree to a satisfactory extent.
Here we investigate the extent of the error made by one such Bayesian phylogenetic tool (BEAST2) when inferring a phylogenetic tree generated by a known species pump diversification model with a standard tree prior.

% Summarising the main result
To this end we simulate our species pump model, which we call the multiple birth model because it produces multiple simultaneous speciation events, under various parameter settings, and evaluate the corresponding error produced during the inference process. We compare this error with the error made when the generating model is the same as the tree prior used in inference.

% Two or three sentences explaining what
% the main result reveals in direct
% comparison to what was thought to be the case
% previously, or how the main result adds to
% previous knowledge.
We show that the extent of the inference error increases with the number of multiple birth events, as expected.
\rampal{We need a bit more reults here}

% One or two sentences to put the results into a
% more general context.
These results allow phylogeneticists to judge under which circumstances the standard tree priors can be safely used, and under what circumstances phylogenetic inferences are unreliable. \giovanni{Not entirely sure if this is the best sentence to put here.}
\rampal{Would be nice if we gave some guidelines here}

### 2020-03-24 razzo-introduction.tex
Feedback rsetienne

Modern computational techniques allow to infer phylogenies from genetic data 
such as DNA, RNA or proteins. BEAST (\cite{beast}) and its descendant 
BEAST2 (\cite{beast2}) are widely used tools to perform such task. They 
return posterior distributions of phylogenies 
and estimated parameters by running a Bayesian analysis, 
given genetic data and tree priors.

A tree prior is a mathematical description 
of the a priori characteristics 
that we expect to be reflected in posterior phylogenies. 
The choice of a specific prior is, by definition, 
arbitrary but the consequences of such choice 
can be vet analyzing the so-obtained posteriors.

BEAST2 gives to the user the option to set up 
several possible phylogenetic 
priors (e.g. substitution/clock/speciation models). 
However, currently available priors 
might be not suitable to analyze some specific datasets.
For this reason BEAST2 provides users with the possibility 
to introduce new tree priors, 
to infer phylogenies based on different assumptions 
on how the speciation process takes place.

Current phylogenetic tools assume that 
only a single speciation event can occur at any given time.
While this assumption has been proved to be useful 
to construct a wide variety of successful 
models (e.g \cite{Maddison2007biSSE}, \cite{Valente2015}, 
\cite{etienne2012diversity}, \cite{etienne2014estimating}), 
they do not take into account the possibility 
for environmentally-driven contemporary speciations on multiple lineages.

We explore such case introducing the multiple-birth death model (MBD), 
currently absent in BEAST2. 
The multiple birth hypothesis aims to describe how 
large-scale environmental changes can lead phylogenies 
with a high abundance of speciation events in relatively short times.

\iffalse
\giovanni{This might be good for the abstract}
The (constant-rate) birth-death (BD) model embodies the common assumption 
that only a single speciation event can occur at any given time. 
The multiple-birth-death (MBD) model relaxes this assumption allowing, 
in addition to standard BD events, 
also events in which large-scale environmental changes lead 
to speciation bursts.
\fi

Such hypothesis can be useful to better understand the history of systems of particular interests for evolutionary biologists, such as the diversification of cichlid fish in the African Great Lakes (Malawi, Tanganyika and Victoria), where water level fluctuations are thought to play an important role in promoting diversification (\citep{verheyen1996mitochondrial}, \citep{sturmbauer2001lake}, \citep{janzen2016}, \citep{janzen2017}).

However, the introduction of new tree priors is not always desirable (\citep{bilderbeek2019pirouette}). Current standard birth death (BD) tree priors might, in principle, prove to be good enough at detecting such events despite their lower level of complexity.
If this is the case one should always choose to adopt the simplest model.

We used the R package \verb;pirouette; (\citep{pirouette}) to perform such test, starting on phylogenies simulated under the MBD regime using the \verb;mbd; package (\citep{mbd}).
From such phylogenies we measure the inference error made adopting a standard BD tree prior in the inference process.

With this work we aim, using such inference error distributions, to test whether or not it is advantageous to implement a new prior model that can allow the construction of trees where multiple speciations can co-occur at the same time.

>>>>

Modern computational techniques, such as BEAST (\cite{beast}) and its descendant 
BEAST2 (\cite{beast2}), allow to infer phylogenetic trees from genetic data 
such as DNA, RNA or proteins. \rampal{mention MrBayes and revBayes as well}. They 
return posterior distributions of phylogenies 
and estimated parameters by running a Bayesian analysis, 
given aligned sequence data and a set of models. 
One of these models is the diversification model, for which a prior distribution must be provided. This is called a tree prior; it is a mathematical description 
of the probability distribution of possible branching patterns before looking at the data. Together with the signal from the data, this tree prior will determine the posterior distribution of phylogenies, i.e. after considering the data. Other models include the nucleotide substitution model (i.e. a model of relative transition rates between different nucleotides through time) and the clock model (a model determining the absolute rate of changes for each lineage). For each of these models choices must be made and prior distributions must be specified for their parameters. BEAST2 gives the user the option to set up 
several possible phylogenetic 
priors (e.g. substitution/clock/diversification models). 
However, currently available priors 
might be not suitable to analyze some specific datasets.
For this reason BEAST2 provides users with the possibility 
to introduce new models and corresponding priors. Particularly, one can specify the tree prior for a new model of diversification.

Current phylogenetic tools such as BEAST2 assume that 
only a single speciation event can occur at any given time.
This assumption is consistent with many different diversification models (e.g \cite{Maddison2007biSSE}, \cite{Valente2015}, 
\cite{etienne2012diversity}, \cite{etienne2014estimating}). However, multiple speciation events can take place simultaneously and repeatedly when populations are intermittently disconnected and connected, for example due to climatic fluctuations. This has been called the species pump hypothesis.\rampal{Cite: Haffer, J. (1969). Speciation in Amazonian forest birds. Science, 165,
131–137. https://doi.org/10.1126/science.165.3889.131.} and has been invoked particularly in mountainous areas that underwent glaciation. \rampal{ Origins of global mountain plant biodiversity: Testing the 'mountain-geobiodiversity hypothesis'
By: Muellner-Riehl, Alexandra N.; Schnitzler, Jan; Kissling, W. Daniel; et al.
JOURNAL OF BIOGEOGRAPHY}. Our own interest in the species pump hypotheses arose from its potential explanation of the radiation of cichlid fish in the African rift lakes (Malawi, Tanganyika and Victoria), where water level drops created multiple smaller lakes providing the opportunity for allopatric speciation in multiple species.  (\citep{verheyen1996mitochondrial}, \citep{sturmbauer2001lake}, \citep{janzen2016}, \citep{janzen2017}).

One could study whether the species pump hypothesis is a viable explanation in empirical ystems by comparing divergence times of sister taxa \rampal{Oaks et al - doi:10.1111/evo.13754}. A more inclusive approach would involve using a model allowing multiple simultaneous speciation events as a new species tree prior in phylogenetic reconstruction. However, introducing a new tree prior may be computationally prohibitive (\citep{bilderbeek2019pirouette}), and may also not be necessary, as current standard birth death (BD) tree priors might prove to be good enough at inferring the correct tree. Here we use the R package \verb;pirouette; (\citep{pirouette}) to check whether this is the case by simulating phylogenies under a species pump model, i.e. the multiple-birth-death model (MBD), with the \verb;mbd; package (\citep{mbd}), then simulating sequence alignments for each of these trees and finally inferring a phylogenetic tree using BEAST2 from these alignment. By comparing the inferred phylogeny with the true (simulated) one, we measure the inference error made by adopting a standard BD tree prior.

### 2020-03-24 razzo-methods.tex
Feedback rsetienne

We proceed in the following way: under a range of parameter setting, we build simulated datasets generated under the multiple birth model. 
Then we run a \verb;pirouette; analysis, which will lead to error distributions between the inferred posterior and the original simulated trees, for each setting. 
Importantly, the \verb;pirouette; analysis includes also a 'twin' parallel pipeline, which will provide a measure of the baseline error due to pure stochasticity, unavoidably occurring in the process.
We then analyse the complete and baseline errors.
\subsection{The Framework}

\subsubsection{Simulation Model}

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the traditional per-species speciation and extinction rates 
already present in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
The first, $\nu$, is the rate at which an environmental change is triggered.

When such event is triggered, each species present in the phylogeny 
at that moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature with respect to 
the one triggered by $\lambda$. 
In fact, whereas parameter $\lambda$ can be seen as 
describing a sympatric process, $\nu$ induces instead the rise of 
geographical barriers able to interrupt the gene flow in the population,
resulting in a possible allopatric speciation for each of the species.  
Even though multiple speciations can co-occur, 
polytomies are not allowed in such process as each species can speciate only 
once during a co-occuring speciation event.

A likelihood expression for the process is provided in \cite{mbd}.

\subsubsection{Estimating the inference error}

From each simulated 'true' MBD tree, we measure the impact of
ignoring the more complex and non-standard MBD tree prior in
Bayesian phylogenetic inference.
We do so by assuming a simpler and standard BD tree prior,
and compare the 'true' tree with the inferred trees.
To do so, we use the \verb;pirouette; R package [\citep{pirouette}].

Here we summarize briefly how \verb;pirouette; works: it starts from a 'true' starting phylogeny (in our case: the simulated MBD tree), from which a DNA sequence alignment is simulated. 
From each sequence alignment, a Bayesian inference is run (in our case: with assuming a BD tree prior), to obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, an inference error distribution is generated.
We use the twinning option available in \verb;pirouette; that allows to quantify the impact of assuming a wrong tree prior in a Bayesian inference compared to a reference background error.

@@

When working on simulated datasets the choice of the generating parameters is always an arbitrary choice. In principle one would choose a parameter range as wide as possible to characterize the most of the available parameter space. However, there are technical aspects that must be taken in consideration to be sure that both the simulation and the inference processes work out in an optimal way.
We required the following criteria to be satisfied \giovanni{Does this apply only to the mbd parameters or also to the pirouette ones?}:
\itemize{
  \item Quality: At least 95\% of all individual runs should have an Estimated Sample Size (ESS) of at least 200 (see \cite{beast} for reference);
    \richel{figure \ref{fig:esses}}
  \item Feasibility: As time is a limited resource we impose that at least 95\% of all individual runs should finish within 10 days;
    \richel{figure \ref{fig:runtimes}}
  \item Reproducibility: the mean run-time of all finished runs should be less than 24 hours \richel{figure \ref{fig:runtimes}}

@@

Additional information about the choice of a suitable parameter setting can be found in the Supplementary Material.
\iffalse
\giovanni{This could be put in the appendix, if we manage to put in a handy format.}
We searched through parameter space until these criteria were met.
All the parameter settings used in the pilot experiments can be found at 
\url{https://github.com/richelbilderbeek/razzo_project/blob/master/overview.md}.
Due to the low number of replicates, we were unable (nor tried)
to draw conclusions based on the results.
\fi

\subsubsection{Simulations setting}

We simulate the speciation process in continuous time using the \verb;mbd_sim; function from the \verb;mbd; R package [\citep{mbd}].
We let parameters vary using all possible combinations of values as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, we generate [as much as time permits] 
\richel{Will need to calculate how many replicates we can do} 
independent trees of the same crown age. Note that trees that are actually passed as \verb;pirouette; input are not full trees, but reconstructed trees instead. A reconstructed tree is a tree where all the species that went extinct before the present time are removed, as this information is usually unavailable on real data phylogenies.

@@

\subsubsection{pirouette setting}

In our experiment, the alignments are 1000 nucleotides in length, with a known root sequence of four 250 mono-nucleotide blocks, generated using the simplest nucleotide substitution model (Jukes Cantor, JC69) and clock model (strict), with a mutation rate of $\frac{t_c}{2}$, where $t_c$ is the crown age. 
With this mutation rate, each nucleotide has a 50\% chance to mutate (both silently and non-silently) from the ancestral root sequence to any of the contemporary species' sequences at the tips.

For the Bayesian inference, we assume a generative model of a JC69 site model, a strict clock model and a BD tree prior.
Additionally, we use a Most Recent Common Ancestor (MRCA) prior equal to the crown age with a normal distribution of width $\sigma = 0.01$. We pick a Markov Chain Monte Carlo (MCMC) setup of $10^6$ states with a sampling interval of once per $10^3$ states. Of the resulting $10^3$ states, we discard a burn-in of $10\%$.

To verify that the generative model (with a BD tree prior) is indeed the most appropriate standard model, we compare it to 39 other candidate inference models.
These candidate models are all other combinations of four tree priors, two clock models and five speciation models, excluding the one combination used as a generative model.
We let the best candidate model also generate an error distribution, which is expected to have a similar or high median.
For models comparison, \verb;pirouette; uses a nested sampling approach to estimate marginal likelihoods, as described in \citet{russel2019model}, using the "\verb;NS;" BEAST2 package. We setup the MCMC for the nested sampling using its default values, which are a relative error $\epsilon$ of $10^{-12}$, 
1 particle, a sub-chain sampling interval of $5 \cdot 10^3$ states with a maximum chain length of $10^6$ states.

To compare the true tree with any of the posterior trees, we use the absolute nLTT statistic by \citet{janzen2015}, which will result in an error distribution with values ranging from zero (when the inferred tree is identical to the true tree) to a maximum of one (trees are completely different).

For the twinning, we let a birth-death process generate one random tree,
so that the twin inference can be done with the correct phylogenetic
model of a JC69 site model, strict clock model and a BD tree prior.

>>>>

\subsection{Simulation model}

The multiple-birth-death (MBD) model inherits the parameters $\lambda$ and $\mu$ from the BD model; they correspond, respectively, 
to the traditional per-species speciation and extinction rates. 
Additionally, the MBD model assumes that external events, occuring at rate $\nu$ triggers a speciation initiation event in each lineage which leads to a full new species with probability $q$. 
Whereas parameter $\lambda$ can be interpreted as the rate of sympatric speciation, $\nu$ is the rate of appearance of 
geographical barriers able to interrupt the gene flow in the population,
resulting in a possible allopatric speciation for each of the species.  
Even though multiple speciation events can occur simultaneouly, it does not lead to  
polytomies, because each species can only split once after a trigger event. This model can be easily simulated with a Doob-Gillespie algorithm. A probability distribution for the phylogeny under the MBD model can also be formulated using the integration approach developed for diversity-dependent diversification models \citep{etienne2012diversity}, which has been explored in \cite{mbd}.\rampal{Is this the mbd chapter?}. While this probability distribution could in principle be used as a tree prior in Bayesian phylogenetic inference, it is computationally very demanding, particularly for large trees. With the MBD model we generate simulated datasets for various parameter settings, using the \verb;mbd_sim; function from the \verb;mbd; R package [\citep{mbd}].

\subsection{Estimating the inference error}

From each simulated 'true' MBD tree, we measure the impact of
ignoring the more complex and non-standard MBD tree prior in
Bayesian phylogenetic inference with the R package \verb;pirouette; [\citep{pirouette}].
 \verb;pirouette; starts from a 'true'  phylogeny (in our case: the simulated MBD tree), and simulates a DNA sequence alignment on it using a known nucleotide substitution model and a clocl model. 
From each sequence alignment, a Bayesian inference is run with a particular choice of tree prior and substitution and clock models. One can choose the same substitution and clock models as used in generating the tree, or pick the ones that fit the data best. For the tree pior we assume the BD model (as the effect of this choice is our focus). We obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, an inference error distribution is generated. For this comparison we used the absolute nLTT statistic by \citet{janzen2015}, which results in an error distribution with values ranging from zero (when the inferred tree is identical to the true tree) to a maximum of one (trees are completely different).

We used the twinning option available in \verb;pirouette; that allows to quantify the impact of assuming a wrong tree prior in a Bayesian inference compared to a reference background error that would arise even if the models used in inference were identical to those used in generating the tree (i.e. the twin tree was generated with a BD model).

@@

When working with simulated datasets the choice of the generating parameters is always somewhat arbitrary. In principle one would choose a parameter range as wide as possible to characterize most of parameter space. However, both simulation and inference need to be possible computationally. To ensure this
we required the following criteria to be satisfied for a parameter set to be chosen: \giovanni{Does this apply only to the mbd parameters or also to the pirouette ones?}:
\itemize{
  \item Quality: At least 95\% of all inferences should have an Estimated Sample Size (ESS) of at least 200 (see \cite{beast} for reference);
    \richel{figure \ref{fig:esses}}
  \item Feasibility: At least 95\% of all individual runs should finish within 10 days;
    \richel{figure \ref{fig:runtimes}}
  \item Reproducibility: the mean run-time of all finished runs should be less than 24 hours \richel{figure \ref{fig:runtimes}}

@@

This resulted in the simulation parameters in Table~\ref{tab:simulation_parameters}.

@@

We assumed that the alignments are 1000 nucleotides in length, with a known root sequence of four 250 mono-nucleotide blocks, generated using the simplest nucleotide substitution model (Jukes Cantor, JC69) and clock model (strict), with a mutation rate of $\frac{t_c}{2}$, where $t_c$ is the crown age. 
With this mutation rate, each nucleotide has a 50\% chance to mutate (both silently and non-silently) from the ancestral root sequence to any of the contemporary species' sequences at the tips.

For the Bayesian inference, we assumed a generative model of a JC69 site model, a strict clock model and a BD tree prior.
Additionally, we used a Most Recent Common Ancestor (MRCA) prior equal to the crown age with a normal distribution of width $\sigma = 0.01$. We used a Markov Chain Monte Carlo (MCMC) setup of $10^6$ states with a sampling interval of once per $10^3$ states. Of the resulting $10^3$ states, we discarded a burn-in of $10\%$.

To verify that the generative model (with a BD tree prior) is indeed the most appropriate standard model, we compared it to 39 other candidate inference models.
These candidate models are all other combinations of four tree priors, two clock models and five speciation models, excluding the one combination used as a generative model.
We let the best candidate model also generate an error distribution, which is expected to have a similar or high median. \rampal{What do you mean by 'high median'?}
For model comparison, \verb;pirouette; uses a nested sampling approach to estimate marginal likelihoods, as described in \citet{russel2019model}, using the "\verb;NS;" BEAST2 package. We setup the MCMC for the nested sampling using its default values, which are a relative error $\epsilon$ of $10^{-12}$, 
1 particle, a sub-chain sampling interval of $5 \cdot 10^3$ states with a maximum chain length of $10^6$ states.

Additional information about the choice of suitable parameter settings can be found in the Supplementary Material.\rampal{What kind of info is this? Explain. Why not put it here?}
\iffalse
\giovanni{This could be put in the appendix, if we manage to put in a handy format.}
We searched through parameter space until these criteria were met.
All the parameter settings used in the pilot experiments can be found at 
\url{https://github.com/richelbilderbeek/razzo_project/blob/master/overview.md}.
Due to the low number of replicates, we were unable (nor tried)
to draw conclusions based on the results.
\fi
